Some examples in extended logic counting classes

### 2017-01-06 LICS/appendix-counting.tex

For the second condition, let $f\in \fpspace$ defined over some $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq 2^{n^\ell}$, where $\A$ has a domain of size $n$. Let $\Phi(X)$ be a $\pfp$ formula that represents the following. Consider an order over the subsets of $A^{\ell}$ given by the formula $$\varphi_{<}(X,Y) = \exists\bar{u}\big[\neg X(\bar{u})\wedge Y(\bar{u})\wedge \forall\bar{v}\big(
\bar{u}<\bar{v}\to(X(\bar{u})\iff Y(\bar{v}))\big)\big].$$

Let $B$ be the $p$-th subset with respect to this order and let $w$ be a string of size $2^{n^{\ell}}$ which represents the value of $f(\enc(\A))$ in binary. Then $\A\models\Phi(B)$ if and only if the $p$-th bit in $w$ is 1, from least to most significant. Since this condition can be verified in polynomial space, and $\pfp$ captures $\pspace$, then $\Phi$ can be expressed in $\pfp$. We define
$\alpha := \sa{X}(\Phi(X)\mult\varphi(X))$, where $\varphi(X) = \pa{Y}(\varphi_{<}(Y,X)\mapsto 2)$. Note that for each $B\subseteq A^{\ell}$ such that $B$ is the $m$-th element in the given order, $\sem{\varphi(B)}(\A) = 2^m$. Therefore, $\alpha\in\qso(\pfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

>>>>

For the second condition, let $f\in \fpspace$ defined over some $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq 2^{n^\ell}$, where $\A$ has a domain of size $n$. Let $X$ be a second-order variable of arity $\ell$. Consider a linear order over predicates of arity $\ell$ given by the formula 
$$
\varphi_{<}(X,Y) = \exists\bar{u}\big[\neg X(\bar{u})\wedge Y(\bar{u})\wedge \forall\bar{v}\big(
\bar{u}<\bar{v}\to(X(\bar{u})\iff Y(\bar{v}))\big)\big].
$$
Consider a polynomial space algorithm over $\R\cup\{X\}$ that receives a $\R\cup\{X\}$-structure $\A'$. Let $X^{\A'}$ be the $p$-th subset in the given order (we consider 0-indexation). The algorithm simulates $f(\enc(\A))$ until $(2^{n^{\ell}}-p)$ characters have been written, but each character of the output is written over the last one. It accepts if and only if the last character written is $1$. Since this algorithm works in polynomial space, it can be described in $\pfp$ \cite{AbiteboulV89} by a formula $\Phi(X)$.

Note that the formula also represents the following.
For each $(\A,v,V)\in\ostr[\R]^*$, let $V(X)$ be the $p$-th subset with respect to this order and let $w$ be a string of size $2^{n^{\ell}}$ which represents the value of $f(\enc(\A))$ in binary. Then $(\A,v,V)\models\Phi(X)$ if and only if the $p$-th bit in $w$ is 1, from least to most significant. We define
$$\alpha := \sa{X}(\Phi(X)\mult\varphi(X)),$$ where $\varphi(X) = \pa{Y}(\varphi_{<}(Y,X)\mapsto 2)$. Note that for each $B\subseteq A^{\ell}$ such that $B$ is the $m$-th element in the given order, $\sem{\varphi(B)}(\A) = 2^m$. Therefore, $\alpha\in\qso(\pfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.
